Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Fixed some refs.

\paragraph{Adapting graph traversal} \ADComment{Tyler, the original had two global barriers; the port has just one resizing barrier.  Is this right?}
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the graph traversal kernel
of \myfig\ref{fig:graphsearch}, introduced in
\mysec\ref{sec:openclexamples}.  
On line~\ref{line:cgraph:resizing}, we change the original global barrier
into a resizing barrier. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barrier
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so that
new workgroups know which frontier they are processing; $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so that new
workgroups know which of the node arrays are input and output. Lastly,
the static work distribution of the original kernel is no longer valid
in a cooperative kernel. This is because the stride (which is based on
$M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}).  This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: \ADComment{Actually I'm confused and need to discuss :-)}

@@

\subsection{Non-Functional Requirements of Cooperative Kernels}\label{sec:nonfunctional}

@@

Using work stealing as an example: in the kernel of \myfig\ref{...}, a
workgroup offers itself to the scheduler after processing each task.
If tasks turn out to be very short then it might turn out to be
reasonable to invoke $\offerkill$ less frequently, after every $k$
tasks for some appropriate $k$.  On the other hand, if individual
tasks may be very time-consuming then it may be necessary to rewrite
the algorithm so work on a larger number of shorter tasks, so as to
achieve a higher rate of calls to $\offerfork$.

Getting this non-functional requirement right depends on the
constraints of the tasks to be co-scheduled with a cooperative kernel,
on properties of the cooperative kernel itself, and on the efficiency
of the GPU that is used for execution.  In \mysec\ref{...} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel in order to
achieve a smooth frame rate, and we evalaute responsiveness in
practice in \mysec\ref{...}.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a comput unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{ }) is as
generous as possible to the cooperative kernel in this regard.
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.

@@

We opted for a semantics whereby only the active workgroup with the
highest id can be killed.  This has an appealing property: it means
that the ids of active workgroups are contiguous, which is important
for processing of contiguous data.  The cooperative graph traversal
algorithm of \myfig\ref{ } illustrates this: the algorithm is prepared
for $\getglobalsize()$ to change after each resizing barrier call, but
depends on the fact that $\getglobalid()$ returns a contiguous range
of thread ids.

\ADComment{Got here.}

@@

\ADComment{Explain here where we could have done things differently.}

\ADComment{What are difficulties with other alternatives and why we didn't use them.}

\ADComment{Allowing zero to be killed; the following is some text we had before:}

\paragraph{Keeping at least one workgroup alive}

@@

\ADComment{Allowing arbitrary workgroups to be killed (not just the largest workgroup)}

\ADComment{Having thread 0 transmit its state, and transmitting only annotated variables.}

@@

\subsection{Determining Representative Workloads}

@@

\subsection{Performance and Responsiveness of Cooperative Scheduling}

>>>>

\paragraph{Adapting graph traversal} \ADComment{Tyler, the original had two global barriers; the port has just one resizing barrier.  Is this right?}
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the graph traversal kernel
of \myfig\ref{fig:graphsearch}, introduced in
\mysec\ref{sec:openclexamples}.  
On line~\ref{line:cgraph:resizing}, we change the original global barrier
into a resizing barrier. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barrier
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so that
new workgroups know which frontier they are processing; $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so that new
workgroups know which of the node arrays are input and output. Lastly,
the static work distribution of the original kernel is no longer valid
in a cooperative kernel. This is because the stride (which is based on
$M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}).  \ADComment{Tyler: can you please add the line number references in the following?} This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: the value of $\keyword{stride}$ obtained from $\getglobalsize()$ at line~\ref{...} is stable until the next resizing barrier at line~\ref{...}, and in particular can be safely used inside the $\keyword{for}$ loop.

@@

\subsection{Non-Functional Requirements}\label{sec:nonfunctional}

@@

Using work stealing as an example: in the kernel of \myfig\ref{fig:workstealing-cooperative}, a
workgroup offers itself to the scheduler after processing each task.
If tasks turn out to be very short then it might turn out to be
reasonable to invoke $\offerkill$ less frequently, after every $k$
tasks for some appropriate $k$.  On the other hand, if individual
tasks may be very time-consuming then it may be necessary to rewrite
the algorithm so work on a larger number of shorter tasks, so as to
achieve a higher rate of calls to $\offerfork$.

Getting this non-functional requirement right depends on the
constraints of the tasks to be co-scheduled with a cooperative kernel,
on properties of the cooperative kernel itself, and on the efficiency
of the GPU that is used for execution.  In \mysec\ref{sec:representativeworkloads} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel in order to
achieve a smooth frame rate, and we evalaute responsiveness in
practice in \mysec\ref{sec:responsiveness}.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a comput unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.

@@

We opted for a semantics whereby only the active workgroup with the
highest id can be killed.  This has an appealing property: it means
that the ids of active workgroups are contiguous, which is important
for processing of contiguous data.  The cooperative graph traversal
algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the algorithm is prepared
for $\getglobalsize()$ to change after each resizing barrier call, but
depends on the fact that $\getglobalid()$ returns a contiguous range
of thread ids.

\ADComment{TODO: Discuss downside, which is that if WG with high id is
  doing a chunky task, a WG with a lower id may be ready to be killed
  but cannot be taken.  Work stealing algorithms, which do not depend
  on thread id, could take advantage of this.}

@@

\ADComment{TODO: To simplify programming, avoiding the need to wrap all kills in conditionals, we always keep one workgroup alive.  This fits well with our decision to kill WGs in decreasing order.}

\ADComment{TODO: There are possible use cases for allowing all WGs to be killed, though, and we might wish to allow them.}

\paragraph{Transmission of partial state from a single thread}

\ADComment{TODO: explain why we have thread 0 transmit its state, and transmitting only annotated variables.}

\ADComment{Something old to maybe work in:}

@@

@@

\subsection{Determining Representative Workloads}\label{sec:representativeworkloads}

@@

\subsection{Performance and Responsiveness of Cooperative Scheduling}\label{sec:responsiveness}
